---
title: "Reconstruct phylogenetic tree of octopamine receptors after creating and exporting tree (in newick format) in MEGA11"
author: "Wee Ye Zhi"
date: "2025-3-17"
output: html_document
---

## Package Installation

# 1.  Install BiocManager, ggtree, treeio, ape, ggplot2, stringr, and dplyr packages
```{r}
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")

BiocManager::install(c("ggtree"))

install.packages(c("ape", "ggplot2", "stringr", "dplyr"))
```

# ===============
# Load Libraries
# ===============

# 2. Load all the libraries
```{r}
library(ggtree)
library(ape)
library(ggplot2)
library(stringr)
library(dplyr)
```

```{r}
packageVersion("ggtree")
packageVersion("ape")
packageVersion("ggplot2")
packageVersion("stringr")
packageVersion("dplyr")
```

```{r}
citation("ggtree")
citation("ape")
citation("ggplot2")
citation("stringr")
citation("dplyr")
```

# ==========
# Read Tree
# ==========

# 3. Import and read the phylogenetic tree (in the form of newick format that is exported from the bootstrap consensus tree) that has been created by MEGA11 previously 
```{r}
tree <- read.tree("C:/Users/USER/Documents/Master_of_Science_in_Systems_Biology_GRA/Experimental_results/Latest_phylogenetic_tree_construction_results/original_octopamine_receptor_phylogenetic_tree.nwk")
```

# ====================
# Define Target Taxon
# ====================

# 4. Define the list of taxon names that you wish to restore their greek letters of alpha and beta
```{r}
original_taxon_name <- c('BmOctalpha1R(NP_001091748.1)', 'CsOctalpha1R(AEQ33589.1)', 'PxOAMB(XP_048486473.1)', 'TcOctalpha1R(NP_001280520.1)', 'DmOctalpha1R(NP_732541.1)', 'AmOctalpha1R(NP_001011565.1)', 'TcOctalpha2R(XP_015839170.1)', 'AmOctalpha2R(XP_001122075.3)', 'CsOctalpha2R(AIC75370.1)', 'DmOctalpha2R(NP_650754.2)', 'PxTyrOctR(XP_048485348.1)', 'CcTyrOctR(TRINITY_DN219751_c0_g1_i1.p1)', 'PrTyrOctR(AFX62896.1)', 'CsTyrOctR(AFG26689.1)', 'BmTyrOctR(BAD11157.1)', 'BmOctbeta2R(NP_001171666.1)', 'PxOctbeta2R(XP_011568733.1)','CcOctbeta2R(TRINITY_DN22425_c0_g1_i3.p1)', 'PrOctbeta2R(XP_022116353.1)','CsOctbeta2R(AEO89318.1)', 'TcOctbeta2R(NP_001280501.1)', 'DmOctbeta2R(NP_001034049.1)', 'NlOctbeta2R(ASA47149.1)', 'AmOctbeta2R(XP_396348.4)', 'TcOctbeta3R(NP_001280505.1)','DmOctbeta3R(NP_001034043.2)', 'AmOctbeta3R(XP_006557730.1)', 'AmOctbeta4R(CCO13925.1)', 'DmOctbeta1R(NP_651057.1)', 'AmOctbeta1R(XP_397139.3)', 'PrOctbeta1R(XP_022123444.1)', 'CsOctbeta1R(AGV79326.1)', 'BmOctbeta1R(XP_004922133.2)', 'VdOctbeta2R(XP_022664697.1)', 'HsOctbeta1AR(NP_000675.1)', 'HsOctbeta2AR(NP_000015.2)', 'HsOctbeta3AR(NP_000016.1)', 'HsOctalpha1AAR(NP_000671.2)', 'HsOctalpha1BAR(NP_000670.1)', 'HsOctalpha1DAR(NP_000669.1)', 'HsOctalpha2AAR(NP_000672.3)', 'HsOctalpha2CAR(NP_000674.2)', 'HsOctalpha2BAR(NP_000673.2)')
```

# =======================
# Restore Greek Letters
# =======================

# 5. Create a 'restore_greek' function to restore the greek letters of alpha and beta for all the alpha and beta receptor subtypes
```{r}
restore_greek <- function(x) {
  x <- str_replace_all(x, "alpha1R", "α1R")
  x <- str_replace_all(x, "alpha2R", "α2R")
  x <- str_replace_all(x, "beta1R", "β1R")
  x <- str_replace_all(x, "beta2R", "β2R")
  x <- str_replace_all(x, "beta3R", "β3R")
  x <- str_replace_all(x, "beta4R", "β4R")
  x <- str_replace_all(x, "beta1AR", "β1AR")
  x <- str_replace_all(x, "beta2AR", "β2AR")
  x <- str_replace_all(x, "beta3AR", "β3AR")
  x <- str_replace_all(x, "alpha1AAR", "α1AAR")
  x <- str_replace_all(x, "alpha1BAR", "α1BAR")
  x <- str_replace_all(x, "alpha1DAR", "α1DAR")
  x <- str_replace_all(x, "alpha2AAR", "α2AAR")
  x <- str_replace_all(x, "alpha2CAR", "α2CAR")
  x <- str_replace_all(x, "alpha2BAR", "α2BAR")
  return(x)
}

tree$tip.label <- restore_greek(tree$tip.label)

tree$tip.label <- str_remove_all(tree$tip.label, "['\"]") # explicitly strip any literal single or double quotes from the labels

tree$tip.label <- str_replace_all(tree$tip.label, " ", "_") # restore underscores (converts spaces back to underscores)
```

# ================================
# Create tip annotation dataframe
# ================================

# 6. Extract the bootstrap values and the tip labels (taxon labels)

```{r}
bootstrap_values <- tree$node.label
restored_taxon_names <- tree$tip.label
```

# 7. Double check whether you have successfully recovered all the Greek letters of alpha and beta, and whether you have annotated protein accession IDs correctly
```{r}
print(restored_taxon_names)
```

# ============================================
# Construct phylogenetic tree for publication
# ============================================

# 8. Plot tree

```{r}
p <- ggtree(tree) +

  geom_tiplab(
    aes(fontface = ifelse(label %in% c(     "CcOctβ2R(TRINITY_DN22425_c0_g1_i3.p1)"), 
                          "bold", "plain")),
    color = "black",    
    size = 3,
    align = TRUE,
    linetype = "solid",
    linesize = 0.3
  ) +

  geom_text2(
    aes(
      subset = !isTip,
      label = label
    ),
    size = 2.5,
    hjust = -0.1,
    vjust = -0.2
  ) +

  theme_tree() +

  coord_cartesian(clip = "off") +

  theme(
    legend.position = "none",   # ← remove legend
    plot.margin = margin(10, 150, 10, 10)
  )

p <- p + hexpand(0.5, direction = 1) # add 50% extra space to the right

print(p)
```

